Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-02-18 draft.tex

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/SNuM.png}
    \caption{SN Ia rate per unit stellar mass as function of galaxy stellar mass. Error bars are the same as Fig. \ref{fig:rate_raw}. SDSS data (orange squares) are taken from \citet{Graur2013}. The systematically higher SNuM in DES is likely caused by the higher redshift of the sample. }%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:snum}
\end{figure}

@@

By binning both the SN hosts and field galaxies by their stellar mass we measure the mean SN Ia rate per galaxy as a function of stellar mass, $R_{\mathrm{G}}(M_*)$. We employ a bootstrap Monte-Carlo approach in order to estimate the uncertainty in the value of the rate in each stellar mass bin. The probability distribution function (PDF) of each galaxy's stellar mass is represented by the sum of two half Gaussian distributions to represent the asymmetric positive and negative uncertainties derived in the SED fitting stage (Section \ref{sec:data}). For each SN host and field galaxy, we take 100 samples of its stellar mass by drawing at random from its PDF. For each of the 100 samples we calculate the SN rate per galaxy in each stellar mass bin via Eq \ref{eq:rate1}, using the completeness-corrected values of $N_{\mathrm{SN}}$ and $N_{\mathrm{SN}}$ from Eqs. \ref{eq:corr_SN} and \ref{eq:corr_G} respectively.

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function in log space, which corresponds to a power law. To find the slope and intercept that best describe the data we use Bayesian inference. To sample the posterior we use the enhanced no-U-turn Sampler (NUTS) algorithm \citep{Betancourt2017}, which is a variant of Hamiltonian Monte-Carlo (HMC), implemented in the Stan programming language \citep{Carpenter2017}. We describe the fitting procedure in full detail in Appendix \ref{appendix:linear_fits}. We measure a slope of $0.70\pm0.02$, which is broadly consistent with the values found in SNLS and SDSS by \citet{Sullivan2006} and \citet{Smith2012} respectively. It is evident that the linear fit is not consistent with the data across the full range of stellar mass considered. In particular, there appear to be two locations of interest: firstly, the slope is steeper at masses lower than $10^9$ M$_{\odot}$; secondly there is an apparent flattening and possibly a turnover at masses around and above $10^{11}$ M$_{\odot}$. The most likely cause of these divergences is because the SN rate is driven by the SFH which is not a linear function of stellar mass (see Section \ref{sec:model}). Second order effects may also be important, and we address these in the following sections.

\subsection{Rate per unit stellar mass \label{subsec:snum}}
\textbf{short section showing SNUM and how it matches past results}

@@

%\textbf{TEMPTED TO DITCH THIS SINCE OUR SFR IS SKETCHY}
%In order to make a more detailed comparison of our results with those of \citet{Sullivan2006} and \citet{Smith2012}, we split galaxies according to their sSFR: galaxies with sSFR$\leq -11.5$ yr$^{-1}$ are denoted as passive, those with $-11.5 < \mathrm{sSFR} \leq -9.5$ are moderately star-forming, and those with sSFR $>-9.5$ are labelled as highly star-forming. The results when split this way are shown in Fig. \ref{fig:rate_raw_split_sfr}. Star-forming galaxies show a steeper slope than their passive counterparts, with values of $1.10 \pm 0.05$ and $0.84 \pm 0.04$ for moderate and highly star-forming galaxies respectively, while passive galaxies show a much shallower slope of $0.55 \pm 0.08$. These values are broadly similar to those measured in SDSS by \citet{Smith2012}, with star-forming galaxies showing steeper slopes than passives (\citet{Smith2012} found 1.01 and 0.72 for SF and passive, respectively). We note that we detect no passive galaxies with mass below $10^{10} \mathrm{M}_{\odot}$. Above this value, the \citet{Smith2012} data for passive galaxies is considerably flatter than that which we measure. On the other hand our results are inconsistent with those found at a similar redshift to the DES data in SLNS \citep{Sullivan2006}, who measured a slope of 1.10 in passives but 0.66 and 0.74 in moderate and highly star-forming galaxies respectively. We note that the moderately star-forming slope is thus consistent between DES and SNLS, but the slopes of passive and highly star-forming galaxies are swapped.
\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results}
	\begin{tabular}{lccr} % four columns, alignment for each
		\hline
		 &$\beta$ & $A$ & $t_{\mathrm{p}}$\\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ & Gyr \\
		\hline
		Fixed $t_{\mathrm{p}}$ & $-0.97_{-0.05}^{+0.05}$ &  $1.73_{-0.05}^{+0.05}$ & -\\
		Free $t_{\mathrm{p}}$ & $-1.02 _{-0.11} ^{+0.08}$ & $1.86 _{-0.13} ^{+0.19}$ & $0.067 _{-0.023} ^{+0.043}$\\
		Fixed $A$, $\beta$ & - & - & $0.047_{-0.007}^{+0.008}$\\
		\hline
	\end{tabular}
\end{table}

@@

By integrating the normalised DTD over cosmic time, we obtain an estimate of the average SN Ia efficiency $N_{\mathrm{Ia}}/M_*$ which represents the number of SNe Ia formed per unit mass of stars formed. We measure $N_{\mathrm{Ia}}/M_* = 2.0~_{-1.6}^{+5.0} \times 10^{-3}~\mathrm{SNe}~\mathrm{M}_{\odot}^{-1}$.

@@

\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results_split_lc}
	\begin{tabular}{crr} % four columns, alignment for each
		\hline
		 Sample &$\beta$ & $A$ \\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ \\
		\hline
		Fiducial & $-1.18\pm0.09$ &  $1.16\pm0.08$ \\
	    $x_1 < 0$ & $-0.79\pm0.08$ &  $1.19 \pm0.05$ \\
		$x_1 > 0$ & $-1.68 _{ -0.15} ^{ +0.14}$ & $0.51\pm0.12$ \\
		$c < 0$ & $-1.09 _{ -0.07} ^{ +0.08}$ & $0.89\pm0.04$ \\
		$c > 0$ & $-1.19\pm0.09$ & $1.16\pm0.08$ \\
		\hline
	\end{tabular}
\end{table}
\subsection{Second order processes affecting the supernova rate \label{subsec:discussion}}

@@

\section{The delay-time distribution as a function of supernova properties}
\label{sec:split_x1_c}

>>>>

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/SNuM.png}
    \caption{SN Ia rate per unit stellar mass as function of galaxy stellar mass. Error bars are the same as Fig. \ref{fig:rate_raw}. SDSS data (orange squares) are taken from \citet{Graur2013}. The systematically higher SNuM in DES is likely caused by the higher redshift of the sample. }%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:snum}
\end{figure}
By binning both the SN hosts and field galaxies by their stellar mass we measure the mean SN Ia rate per galaxy as a function of stellar mass, $R_{\mathrm{G}}(M_*)$. We employ a bootstrap Monte-Carlo approach in order to estimate the uncertainty in the value of the rate in each stellar mass bin. The probability distribution function (PDF) of each galaxy's stellar mass is represented by the sum of two half Gaussian distributions to represent the asymmetric positive and negative uncertainties derived in the SED fitting stage (Section \ref{sec:data}). For each SN host and field galaxy, we take 100 samples of its stellar mass by drawing at random from its PDF. For each of the 100 samples we calculate the SN rate per galaxy in each stellar mass bin via Eq \ref{eq:rate1}, using the completeness-corrected values of $N_{\mathrm{SN}}$ and $N_{\mathrm{SN}}$ from Eqs. \ref{eq:corr_SN} and \ref{eq:corr_G} respectively.

The rate of SNe Ia per galaxy as a function of stellar mass is shown in Fig. \ref{fig:rate_raw}. The relationship between SN Ia rate and galaxy stellar mass is well described by a linear function in log space, which corresponds to a power law. To find the slope and intercept that best describe the data we use Bayesian inference. To sample the posterior we use the enhanced no-U-turn Sampler (NUTS) algorithm \citep{Betancourt2017}, which is a variant of Hamiltonian Monte-Carlo (HMC), implemented in the Stan programming language \citep{Carpenter2017}. We describe the fitting procedure in full detail in Appendix \ref{appendix:linear_fits}. We measure a slope of $0.63\pm0.02$, which is consistent with the value found in weakly star-forming galaxies in SNLS by \citet{Sullivan2006}, but also in passive galaxies in SDSS by \citet{Smith2012}, and in all galaxies in LOSS by \citet{Li2011a}.

A best fitting slope of less than unity (in log space) indicates a non-linear relationship between SN Ia rate and galaxy stellar mass. Indeed, the rate is instead likely to be driven by the DTD which is likely a non-linear function of stellar age. Investigating the shape of this DTD forms the basis for the latter part of this paper in Sections \ref{sec:model} and \ref{sec:split_x1_c}.

@@

%\textbf{TEMPTED TO DITCH THIS SINCE OUR SFR IS SKETCHY}
%In order to make a more detailed comparison of our results with those of \citet{Sullivan2006} and \citet{Smith2012}, we split galaxies according to their sSFR: galaxies with sSFR$\leq -11.5$ yr$^{-1}$ are denoted as passive, those with $-11.5 < \mathrm{sSFR} \leq -9.5$ are moderately star-forming, and those with sSFR $>-9.5$ are labelled as highly star-forming. The results when split this way are shown in Fig. \ref{fig:rate_raw_split_sfr}. Star-forming galaxies show a steeper slope than their passive counterparts, with values of $1.10 \pm 0.05$ and $0.84 \pm 0.04$ for moderate and highly star-forming galaxies respectively, while passive galaxies show a much shallower slope of $0.55 \pm 0.08$. These values are broadly similar to those measured in SDSS by \citet{Smith2012}, with star-forming galaxies showing steeper slopes than passives (\citet{Smith2012} found 1.01 and 0.72 for SF and passive, respectively). We note that we detect no passive galaxies with mass below $10^{10} \mathrm{M}_{\odot}$. Above this value, the \citet{Smith2012} data for passive galaxies is considerably flatter than that which we measure. On the other hand our results are inconsistent with those found at a similar redshift to the DES data in SLNS \citep{Sullivan2006}, who measured a slope of 1.10 in passives but 0.66 and 0.74 in moderate and highly star-forming galaxies respectively. We note that the moderately star-forming slope is thus consistent between DES and SNLS, but the slopes of passive and highly star-forming galaxies are swapped.
\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results}
	\begin{tabular}{lccr} % four columns, alignment for each
		\hline
		 &$\beta$ & $A$ & $t_{\mathrm{p}}$\\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ & Gyr \\
		\hline
		Fixed $t_{\mathrm{p}}$ & $-1.13\pm0.5$ &  $2.08 \pm 0.08$ & -\\
		Free $t_{\mathrm{p}}$ & $-1.02 _{-0.11} ^{+0.08}$ & $1.86 _{-0.13} ^{+0.19}$ & $0.067 _{-0.023} ^{+0.043}$\\
		Fixed $A$, $\beta$ & - & - & $0.047_{-0.007}^{+0.008}$\\
		\hline
	\end{tabular}
\end{table}

@@

By integrating the normalised DTD over cosmic time, we obtain an estimate of the average SN Ia efficiency $N_{\mathrm{Ia}}/M_*$ which represents the number of SNe Ia formed per unit mass of stars formed. We measure $N_{\mathrm{Ia}}/M_* = 0.9~_{-0.7}^{+4.0} \times 10^{-3}~\mathrm{SNe}~\mathrm{M}_{\odot}^{-1}$, which is consistent with \citet{Graur2011}, \citet{Maoz2011}, \citet{Perrets2012}, and \citet{Graur2013}

@@

\renewcommand{\arraystretch}{1.2}
\begin{table}
	\centering
	\caption{Results of the Bayesian parameter estimation for the SN Ia DTD.}
	\label{tab:dtd_results_split_lc}
	\begin{tabular}{crr} % four columns, alignment for each
		\hline
		 Sample &$\beta$ & $A$ \\
		 &-       & $10^{-13}$ M$_{\odot}^{-1}$yr$^{-1}$ \\
		\hline
		Fiducial & $-1.14\pm0.05$ &  $2.08\pm0.08$ \\
	    $x_1 < 0$ & $-0.79\pm0.08$ &  $1.19 \pm0.05$ \\
		$x_1 > 0$ & $-1.68 _{ -0.15} ^{ +0.14}$ & $0.51\pm0.12$ \\
		$c < 0$ & $-1.09 _{ -0.07} ^{ +0.08}$ & $0.89\pm0.04$ \\
		$c > 0$ & $-1.19\pm0.09$ & $1.16\pm0.08$ \\
		\hline
	\end{tabular}
\end{table}
\subsection{Second order processes affecting the supernova rate \label{subsec:discussion}}

@@

\subsection{Rate per unit stellar mass \label{subsec:snum}}

An alternative view on how the rate of SNe relates to environment can be gained by calculated the rate of SNe per year per unit stellar mass: SNuM. A measurement of the SNuM is achieved by summing the number of SNe in each stellar mass bin, and dividing by the total stellar mass of field galaxies in that bin. We show the SNuM as a function of galaxy stellar mass in Fig.~ \ref{fig:snum}, along with the measurements from SNe discovered in SDSS spectra \citep{Graur2013} at $z\sim0.1$. We also calculate the SNuM predicuted by our model for each simulated galaxy at a range of redshifts and plot these as curves in Fig.~\ref{fig:snum}. Our model accurately recovers the SNuM measured in DES for the redshift and stellar mass range of the data. While the data from \citet{Graur2013} lies slightly higher than our model prediction, the difference is consistent with differences in assumed SFHs and stellar population templates as well as SN detection method.
\section{The delay-time distribution as a function of supernova properties}
\label{sec:split_x1_c}
